\documentclass{beamer}
\usepackage{natbib}
\usepackage{amsfonts}
\usepackage{amsmath}
\usepackage{mathtools}
\usepackage{dsfont}
\usepackage{algorithm}
\usepackage{algorithmic}

\DeclareMathOperator*{\argmax}{arg\,max}
\DeclareMathOperator{\diag}{diag}

\usetheme{CambridgeUS}

\title{Applications of Hidden Markov Models}
\author{Andrew Floren}

\begin{document}

\begin{frame}
\titlepage
\end{frame}

\section*{Introduction}
\subsection*{Hidden Markov Model}
\begin{frame}{Introduction}
Let $X_1, X_2, \cdots, X_t$ be a Markov chain on the discrete domain $\Omega$ with transition probability matrix $P \in \mathbb{R}^{|\Omega| \times |\Omega|} $ where $P_{i,j} = P(X_{t+1} = j | X_{t} = i)$. Let $Y_1, Y_2, \cdots, Y_t$ be a series of random variables on the discrete domain $\Phi$ with observation probability matrix $O \in \mathbb{R}^{|\Omega| \times |\Phi|}$ where $O_{i,j} = P(Y_t = j | X_t = i)$ and each $Y_t$ is conditionally independent given $X_t$.
\end{frame}

\section{Filtering}
\begin{frame}{Filtering}
Want to find
\begin{equation}
\label{eqn:filtering}
P(X_t | Y_1, Y_2, \cdots, Y_t).
\end{equation}
\end{frame}

\begin{frame}
First, let us consider how to calculate the full joint probability 
\begin{equation}
P(X_{t}, Y_{1}, Y_{2}, \cdots, Y_t)
\end{equation}
\end{frame}

\begin{frame}
\begin{align}
&P(X_{t}, Y_{1}, Y_{2}, \cdots, Y_t)  \label{eqn:forward-message}\\
&= P(Y_{t} | X_{t}, Y_{1}, \cdots, Y_{t-1}) P(X_{t}, Y_{1}, \cdots, Y_{t-1}) \\ 
&= P(Y_{t} | X_{t}) P(X_{t}, Y_{1}, \cdots, Y_{t-1}) \\
&= P(Y_{t} | X_{t}) \sum_{X_{t-1} \in \Omega}{P(X_{t}, X_{t-1}, Y_{1}, \cdots, Y_{t-1})} \\
&= P(Y_{t} | X_{t}) \sum_{X_{t-1} \in \Omega}{P(X_{t} | X_{t-1}, Y_{1}, \cdots, Y_{t-1}) P(X_{t-1}, Y_{1}, \cdots, Y_{t-1})} \\
&= P(Y_{t} | X_{t}) \sum_{X_{t-1} \in \Omega}{P(X_{t} | X_{t-1})P(X_{t-1}, Y_{1}, \cdots, Y_{t-1})}
\end{align}
Note that $P(Y_{t} | X_{t})$ and $P(X_{t} | X_{t-1})$ are defined directly by our model, and $P(X_{t-1}, Y_{1}, \cdots, Y_{t-1})$ is the same as equation \ref{eqn:forward-message} at time $t-1$.
\end{frame}

\begin{frame}
To recover the conditional probability we calculate
\begin{equation}
P(X_t | Y_1, Y_{2}, \cdots, Y_t) = \frac{P(X_t, Y_1, Y_{2}, \cdots, Y_t)}{\sum_{X_t \in \Omega}{P(X_t, Y_1, Y_{2}, \cdots, Y_t)}}.
\end{equation}
\end{frame}

\subsection{Forward Algorithm}
\begin{frame}{Forward Algorithm}
Given observations $y_1, y_2, \cdots, y_t$, a probability transition matrix $P$, an observation probability matrix $O$, an initial state distribution $\pi$, and $\alpha_i(x) = P(X_i = x, Y_1 = y_1, \cdots, Y_i = y_i)$, the filtered distribution of state $X_t \in \Omega$ can be determined using Algorithm \ref{alg:forward}.
\begin{algorithm}[H]
\caption{Forward Algorithm}
\label{alg:forward}
\begin{algorithmic}
\STATE $\alpha_0 \leftarrow \pi$
\FOR{$i \leftarrow 1$ \TO $t$}
\STATE $\alpha_i \leftarrow \diag(O_{*,y_t}) \cdot P \cdot \alpha_{i-1}$
\ENDFOR
\RETURN $\frac{\alpha_t}{\|\alpha_t\|_1}$
\end{algorithmic}
\end{algorithm}
\end{frame}

\section{Smoothing}
\begin{frame}{Smoothing}
Want to find
\begin{equation}
\label{eqn:smoothing}
P(X_k | Y_1, Y_2, \cdots, Y_t), 
\end{equation}
where $0<k<t$.
\end{frame}

\begin{frame}
Similar to filtering, let us begin by considering the full joint probability 
\begin{equation}
P(X_{k}, Y_{1}, Y_{2}, \cdots, Y_t).
\end{equation}
\end{frame}

\begin{frame}
\begin{align}
&P(X_k, Y_1, \cdots, Y_t) \\
&= P(Y_{k+1}, \cdots, Y_t | X_k, Y_1, \cdots, Y_k) P(X_k,  Y_1, \cdots, Y_k) \\
&= P(Y_{k+1}, \cdots, Y_t | X_k) P(X_k,  Y_1, \cdots, Y_k)
\end{align}
Note that $P(X_k, Y_1, \cdots, Y_k)$ is the unnormalized output of the forward algorithm up to time $k$. Therefore, we only need to solve 
\begin{equation}
P(Y_{k+1}, \cdots, Y_t | X_k).
\end{equation}
\end{frame}

\begin{frame}
\begin{align}
&P(Y_{k+1}, \cdots, Y_t | X_k) \label{eqn:backward-message}\\
&= \sum_{X_{k+1} \in \Omega}{P(Y_{k+1}, \cdots, Y_t, X_{k+1} | X_k)} \\
&= \sum_{X_{k+1} \in \Omega}{P(Y_{k+1}, \cdots, Y_t | X_{k+1} , X_k) P(X_{k+1} | X_k)} \\
&= \sum_{X_{k+1} \in \Omega}{P(Y_{k+1}, \cdots, Y_t | X_{k+1}) P(X_{k+1} | X_k)} \\
&= \sum_{X_{k+1} \in \Omega}{P(Y_{k+1} | X_{k+1}) P(Y_{k+2}, \cdots, Y_t | X_{k+1}) P(X_{k+1} | X_k)} 
\end{align}
$P(Y_{k+1} | X_{k+1})$ and $P(X_{k+1} | X_k)$ are defined directly by our model, and $P(Y_{k+2}, \cdots, Y_t | X_{k+1})$ is the same as equation \ref{eqn:backward-message} at time $k+1$.
\end{frame}

\begin{frame}
To recover the conditional probability we calculate
\begin{equation}
P(X_k | Y_1, Y_{2}, \cdots, Y_t) = \frac{P(X_k, Y_1, Y_{2}, \cdots, Y_t)}{\sum_{X_k \in \Omega}{P(X_k, Y_1, Y_{2}, \cdots, Y_t)}}
\end{equation}
\end{frame}

\subsection{Forward-Backward Algorithm}
\begin{frame}{Forward-Backward Algorithm}
Given $y_1, y_2, \cdots, y_t$, $P$, $O$, $\pi$, $\alpha_i(x) = P(X_i = x , Y_1 = y_1, \cdots, Y_i = y_i)$, and $\beta_i(x) = P(Y_{i+1} = y_{i+1}, \cdots, Y_i = y_i | X_i = x)$, the smoothed distribution of state $X_k \in \Omega$ can be determined using Algorithm \ref{alg:forward-backward}.
\begin{algorithm}[H]
\caption{Forward-Backward Algorithm}
\label{alg:forward-backward}
\begin{algorithmic}
\STATE $\alpha_0 \leftarrow \pi$
\FOR{$i \leftarrow 1$ \TO $k$}
\STATE $\alpha_i \leftarrow \diag(O_{*,y_i}) \cdot P \cdot \alpha_{i-1}$
\ENDFOR
\STATE $\beta_t \leftarrow \mathds{1}$
\FOR{$i \leftarrow t-1$ \TO $k$}
\STATE $\beta_i \leftarrow \diag(O_{*,y_{i+1}}) \cdot P \cdot \beta_{i+1}$
\ENDFOR
\RETURN $\frac{\alpha_k \cdot \beta_k}{\|\alpha_k \cdot \beta_k\|_1}$
\end{algorithmic}
\end{algorithm}
\end{frame}

\section{Maximum Likelihood Sequence}
\begin{frame}{Maximum Likelihood Sequence}
Want to find 
\begin{equation}
\argmax_{x_1, x_2, \cdots, x_t \in \Omega} P(X_1 = x_1, X_2 = x_2, \cdots, X_t = x_t | Y_1, Y_2, \cdots, Y_t).
\end{equation}
\end{frame}

\begin{frame}
Consider
\begin{align}
&\max_{x_1, \cdots, x_t} P(X_1 = x_1, \cdots, X_t = x_t, X_{t+1}, Y_1, \cdots, Y_t, Y_{t+1})&\\
&\begin{multlined}
= P(Y_{t+1} | X_{t+1}) \max_{X_t \in \Omega}(\\
P(X_{t+1} | X_t) \max_{x_1, \cdots, x_{t-1} \in \Omega}P(X_1 = x_1, \cdots, X_{t-1} = x_{t-1}, X_t,\\
Y_1, \cdots, Y_{t-1}, Y_t)).
\end{multlined}
\end{align}
Note the strong similarity to the filtering problem except we are taking the max across states rather than the sum.
\end{frame}

\subsection{Viterbi Algorithm}
\begin{frame}{Viterbi Algorithm}
Given $y_1, y_2, \cdots, y_t \in \Phi$, $P$, $O$, $\pi$, and $\alpha_t(x) = \max_{x_1, \cdots, x_{t-1} \in \Omega}P(X_1 = x_1, \cdots, X_{t-1} = x_{t-1}, X_t = x, Y_1 = y_1, \cdots, Y_{t-1} = y_{t-1}, Y_t = y_t)$, we can determine the maximum probability sequence $x_1, x_2, \cdots, x_t \in \Omega$ using Algorithm \ref{alg:viterbi}.
\begin{algorithm}[H]
\caption{The Viterbi Algorithm}
\label{alg:viterbi}
\begin{algorithmic}
\STATE $\alpha_0 \leftarrow \pi$
\FOR{$i \leftarrow 1$ \TO $t$}
\FORALL{$x$ such that $x \in \Omega$}
\STATE $\alpha_i(x) \leftarrow O_{x,y_i} \cdot \max{\left( \diag(P_{x,*}) \cdot \alpha_{i-1} \right)}$
\ENDFOR
\STATE $x_i \leftarrow \argmax_{x \in \Omega}{\alpha_i(x)}$
\ENDFOR
\RETURN $(x_1, \cdots, x_t)$
\end{algorithmic}
\end{algorithm}
\end{frame}

\section{Decoding fMRI}
\begin{frame}{Decoding fMRI}
In my research, I am attempting to decode brain states from functional magnetic resonance imaging (fMRI) data.
The level of neural activation in the brain is approximately captured by the blood-oxygen-level dependent (BOLD) signal.
However, this signal is extremely noisy; the signal-to-noise ratio is generally on the order of 1.
Therefore, we must integrate repeated measures to achieve good accuracy.
Traditionally, this is done by averaging across blocks of time where the stimulus, and presumably the brain state, is constant.
By leveraging a hidden Markov model, we can relax the constraints on the temporal structure of the stimulus and hidden brain states.
However, to accurately model the state transition probabilities we need to use a discrete number of previous states.
\end{frame}

\subsection{Model}
\begin{frame}{Model}
The set of brain states we are interested in decoding is $\omega$, where $|\omega| = 6$ for our stimulus.
The domain of the hidden states $X_1, \cdots, X_t$ is $\Omega = \omega^k$, where $k = 6$ is the number of memory states we use.
This results in an extremely large but sparse state transition matrix $P$.
The domain of the observable sequence $Y_1, \cdots, Y_t$ is $\Phi = \mathbb{R}^n$ where $n$ is the number of voxels we are measuring.
We use a trained feed-forward neural network to approximate $P(X_t | Y_t)$ from the observed fMRI data.
We can't generate an observation probability matrix because our observable domain is continuous, but by fitting a distribution to $P(Y_t)$ we can calculate $P(Y_t | X_t)$ using Bayes' rule.
Fortunately, prior fMRI studies have shown that the distribution of fMRI data can be reasonably approximated by a multi-variate Gaussian after whitening.
\end{frame}

\subsection{Modified Forward Algorithm}
\begin{frame}{Modified Forward Algorithm}
Given observations $y_1, y_2, \cdots, y_t \in \Phi$, a probability transition matrix $P$, an observation probability distribution $P_y(x) \sim \mathbb{N}(\mu,\sigma)$, a function $N: \Phi \to \Omega$ that approximates $P(X_t | Y_t = y_t)$, an initial state distribution $\pi$, and $\alpha_i(x) = P(X_i = x, Y_1 = y_1, \cdots, Y_i = y_i)$, the filtered distribution for the state $X_t \in \Omega$ can be determined using Algorithm \ref{alg:modified-forward}.
The performance of this algorithm should be estimated using a cross-validation procedure because the distribution $P_y$ and the function $N$ need to be constructed on a training set.
\begin{algorithm}[H]
\caption{Modified Forward Algorithm}
\label{alg:modified-forward}
\begin{algorithmic}
\STATE $\alpha_0 \leftarrow \pi$
\FOR{$i \leftarrow 1$ \TO $t$}
\STATE $\alpha_i(x) \leftarrow \frac{N(y_t) P_y(y_t)}{\pi} \cdot P \cdot \alpha_{i-1}$
\ENDFOR
\RETURN $\frac{\alpha_t}{\|\alpha_t\|_1}$
\end{algorithmic}
\end{algorithm}
\end{frame}

\subsection{Results}
\begin{frame}{Results}
Still processing...
\end{frame}

\bibliographystyle{IEEEtranN}
\nocite{Russell:2009:AIM:1671238}
\nocite{viterbi1967error}
\nocite{Richard1991}
\begin{frame}{References}
\bibliography{bibliography}
\end{frame}

\end{document}